Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2022-12-22 ms/ms03_results.tex
MS

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
% TODO Supplementary table 1 for newest pipeline
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \citep{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
% TODO Supplementary table 2 for newest pipeline
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
Among these studies, there were 8,580 clumped leading variants with a p-value below 5e-8 from 284 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits; 284
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 8,580
% tophits goes to OSF
% colocpleio goes to OSF

My pipeline uncovered eQTL colocalizations for 5,484 or 64$\%$ of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.rsid from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75) as sel; 5484
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127

The largest classes of diseases by number of leading GWAS variants are autoimmune diseases (2107), cancer of breast (1976) and allergy (1321)
%
In these diseases, the percentages of variants with a colocalized eQTL are 37\%, 21\% and 42\% in these classe, respectively (Supplementary table ST3).
% TODO
% Number of tophits rsid per GWAS class
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits where tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc;
% Number of tophits rsid per GWAS class that is also preset in coloc table
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits, coloc where tophits.gwas_id=gwas_annot.gwas_id and tophits.rsid=coloc.rsid group by gwas_class order by count desc;
% OLD SQLITE
% select count(distinct tophits.rsid),gwas_class from gwas_annot join tophits on tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc limit 5;
%/* count tophits.rsid group by gwas_annot.gwas_class where coloc.`PP.H4.abf`>=0.8*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8 GROUP by gwas_annot.gwas_class;
%/* count tophits.rsid group by gwas_annot.gwas_class*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id GROUP by gwas_annot.gwas_class order by gwas_annot.gwas_class;

These results are consistent with previous colocalization analysis that found explained percentages of work for autoimmune diseases \citep{2021.Li.Mu}.....
% TODO
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

This pipeline resulted in 103,551 variants from 246 GWAS that colocalized (PP.H4.abf$\geq$0.8) with at least one eQTL.
% TODO
%select count(coloc.rsid) from (select DISTINCT gwas_id FROM coloc where pp_h4_abf>=0.8);
%select count(*) from (select distinct rsid from coloc where pp_h4_abf>=0.8) as sel;
%
This might be a useful resource to annotate GWAS variants with genes and biosamples.
Therefore I have developed a web application where I expose 116,736??? variants from 246??? GWAS that colocalize with at least one eQTL at a threshold value PP.H4.abf$\geq$0.75.
% TODO
% select count(*) from (select DISTINCT chrom,pos38,alt FROM colocweb) as sel;
% select count(*) from (select DISTINCT gwas_id FROM coloc) as sel;

@@

% TODO Check if I aggregate different coloc windows or not.
I also computed trait distances based on the spearman correlation of eQTL effect size for distinct eQTLs given by
distinct chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
From this point on I defined colocalization with parameters PP.H4.abf>=0.75 and SNP.PP.H4>=0.5.
%
The argument for selecting a cutoff SNP.PP.H4>=0.5 is given below.
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
I plotted only trait classes with more than a minimum number of traits (Fig. \ref{fig:gwas_distance}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:gwas_distance}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

% TODO create supplementary table 2
I manually assigned the 418 GWAS to 112 classes to aggregate identical or similar traits (Supplementary table 2).

@@

These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The most pleiotropic variants belong to 6 classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in cytobands 1p13.2, 12q24.12 and 19.q13.33.
%
The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%
Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC

>>>>

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \citep{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
These studies resulted in 8,580 clumped leading variants with a p-value below 5e-8 from 284 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits; 284
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 8,580
% tophits goes to OSF
% colocpleio goes to OSF

My pipeline uncovered eQTL colocalizations for 5,484 or 64$\%$ of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75) as sel; 5484
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127

The largest classes of diseases by number of leading GWAS variants are cancer of breast (1976), allergy (1288), and autoimmune diseases (1216).
%
In these disease classes, the percentages of variants with a colocalized eQTL are 58\%, 61\% and 68\%, respectively (Supplementary table ST3).
% Supplementary ST3
% Number of tophits rsid per GWAS class
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits where tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc;
% Number of tophits rsid per GWAS class that is also preset in coloc table
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits, coloc where tophits.gwas_id=gwas_annot.gwas_id and tophits.rsid=coloc.rsid group by gwas_class order by count desc;

These results is much higher than previous colocalization analyses for autoimmune diseases by previous authors that
found 23-41\% depending on the study \citep{2016.Soranzo.Chen,2017.Cotsapas.Chun,2021.Li.Mu}.
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

These data is a useful resource to annotate GWAS variants with genes and biosamples.
Therefore I have developed a web application where I expose 138,301 variants from 293 GWAS that colocalize with at least
one eQTL at a threshold value PP.H4.abf$\geq$0.75.
% TODO
% select count(*) from (select DISTINCT chrom,pos38,alt FROM colocweb) as sel; 138301
% select count(*) from (select DISTINCT gwas_id FROM colocweb) as sel; 293

@@

Then I wanted to verify if different traits cluster coherently in different trait classes.
%
As trait distance, I used the the spearman correlation of the eQTL splitted by
chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
To avoid non resolve eQTLs, I set a cutoff SNP.PP.H4>=0.5 in addition to the cutoff PP.H4.abf>=0.75.
%
The cutoff SNP.PP.H4>=0.5 was chosen, it allow to group pleiotropic eQTLs in a reasonable number of regions (See below).
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
Only trait classes with more than a minimum number of traits were plotted (Fig. \ref{fig:gwas_distance}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
In the OpenGWAS, some traits are redundant because the come from the same reference.
%
I plotted non-rendundant GWAS traits.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:gwas_distance}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

I manually assigned the 418 GWAS to 112 classes to aggregate identical or similar traits (Supplementary table 2).

@@

% TODO CONTINUE
These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The most pleiotropic variants belong to 6 classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in cytobands 1p13.2, 12q24.12 and 19.q13.33.
%
The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%
Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC
